### 2017-05-12 paper/arrows.tex
spacing in figure enumerations.

instance Monad m => Arrow (Kleisli m) where
	arr f = Kleisli (return . f)
	f >>> g = Kleisli (\a -> f a >>= g)
	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
\end{lstlisting}%$
\caption{Definition of the Kleisli type and the corresponding arrow instance}
\label{fig:arrowkleisli}
\end{figure}
\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
		\subcaptionbox{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
		\subcaptionbox{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		\subcaptionbox{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
	\end{tabular}
	\caption{Syntactic sugar for arrows}
	\label{3}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second} (Fig.~\ref{fig:secondImg},\ref{fig:second}),
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
second :: Arrow arr => arr a b -> arr (c, a) (c, b)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{lstlisting}
\caption{The second combinator}
\label{fig:second}
\end{figure}
the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},\ref{fig:***})
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{lstlisting}
\caption{The (***) combinator}
\label{fig:***}
\end{figure}
and the \code{\&\&\&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input (Fig.~\ref{fig:&&&Img},\ref{fig:&&&}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{lstlisting}
\caption{The (\&\&\&) combinator}
\label{fig:&&&}
\end{figure}
A short example given by Hughes on how to use this is \lstinline{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{lstlisting}
\caption{Add over arrows}
\label{fig:addArrows}
\end{figure}
% The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
% \\\\
% \textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used. We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}
%

>>>>

instance Monad m => Arrow (Kleisli m) where
	arr f = Kleisli (return . f)
	f >>> g = Kleisli (\a -> f a >>= g)
	first f = Kleisli (\(a,c) -> f a >>= \b -> return (b,c))
\end{lstlisting}%$
\caption{Definition of the Kleisli type and the corresponding arrow instance}
\label{fig:arrowkleisli}
\end{figure}
\begin{figure}[h]
	\centering
	\begin{tabular}{cc}
		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
		\subcaptionbox{second\label{fig:secondImg}}{\includegraphics[width = 1.5in]{images/second}} \\
		\subcaptionbox{(***)\label{fig:***Img}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		\subcaptionbox{(\&\&\&)\label{fig:&&&Img}}{\includegraphics[width = 1.5in]{images/andandand}}\
	\end{tabular}
	\caption{Syntactic sugar for arrows}
	\label{3}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \code{first}, called \code{second} (Fig.~\ref{fig:secondImg},~\ref{fig:second}),
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
second :: Arrow arr => arr a b -> arr (c, a) (c, b)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{lstlisting}
\caption{The second combinator}
\label{fig:second}
\end{figure}
the *** combinator which combines first and second to handle two inputs in one arrow, (Fig.\ref{fig:***Img},~\ref{fig:***})
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{lstlisting}
\caption{The (***) combinator}
\label{fig:***}
\end{figure}
and the \code{\&\&\&} combinator that constructs an arrow which outputs two different values like \code{***}, but takes only one input (Fig.~\ref{fig:&&&Img},~\ref{fig:&&&}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{lstlisting}
\caption{The (\&\&\&) combinator}
\label{fig:&&&}
\end{figure}
A short example given by Hughes on how to use this is \lstinline{add} over arrows, which can be seen in Fig.~\ref{fig:addArrows}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{lstlisting}
\caption{Add over arrows}
\label{fig:addArrows}
\end{figure}
% The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
% \\\\
% \textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used. We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}
%

### 2017-05-12 paper/parrows.tex
spacing in figure enumerations.

\section{Parallel Arrows}
\label{sec:parallel-arrows}
We have seen what Arrows are and how they can be used as a general interface to computation. In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well. We start by introducing the interface and explaining the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\subsection{The ArrowParallel typeclass}
As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},\ref{fig:parEvalN}).
Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \lstinline{listApp} from Appendix \ref{utilfns}, but does parallel instead of serial evaluation).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{lstlisting}
\caption{parEvalN Arrow combinator as a function}
\label{fig:parEvalNArrowFn}
\end{figure}
With this definition of \code{parEvalN}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we introduce a new typeclass \code{ArrowParallel arr a b} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
\caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
\label{fig:parEvalNArrowTypeClass1}
\end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs. normalform). For this reason we also introduce an additional \code{conf} parameter to the function. We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}. So we add it to the type signature of the typeclass as well and get \code{ArrowParallel arr a b conf} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{lstlisting}
\caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
\label{fig:parEvalNArrowTypeClassFinal}
\end{figure}
Note that we don't require the \code{conf} parameter in every implementation. If it is not needed, we usually just default the \code{conf} type parameter to () and even blank it out in the parameter list of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.

>>>>

\section{Parallel Arrows}
\label{sec:parallel-arrows}
We have seen what Arrows are and how they can be used as a general interface to computation. In the following section we will discuss how Arrows constitute a general interface not only to computation, but to \textbf{parallel computation} as well. We start by introducing the interface and explaining the reasonings behind it. Then, we discuss some implementations using exisiting parallel Haskells. Finally, we explain why using Arrows for expressing parallelism is beneficial.
\subsection{The ArrowParallel typeclass}
As we have seen earlier, in its purest form, parallel computation (on functions) can be seen as the execution of some functions \code{a -> b} in parallel, \code{parEvalN} (Fig.~\ref{fig:parEvalNTypeSig},~\ref{fig:parEvalN}).
Translating this into arrow terms gives us a new operator \code{parEvalN} that lifts a list of arrows \code{[arr a b]} to a parallel arrow \code{arr [a] [b]} (Fig.~\ref{fig:parEvalNArrowFn}) (This combinator is similar to our utility function \lstinline{listApp} from Appendix \ref{utilfns}, but does parallel instead of serial evaluation).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
parEvalN :: (Arrow arr) => [arr a b] -> arr [a] [b]
\end{lstlisting}
\caption{parEvalN Arrow combinator as a function}
\label{fig:parEvalNArrowFn}
\end{figure}
With this definition of \code{parEvalN}, parallel execution is yet another arrow combinator. But as the implementation may differ depending on the actual type of the arrow \code{arr} and we want this to be an interface for different backends, we introduce a new typeclass \code{ArrowParallel arr a b} to host this combinator (Fig.~\ref{fig:parEvalNArrowTypeClass1}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b where
	parEvalN :: [arr a b] -> arr [a] [b]
\end{lstlisting}
\caption{parEvalN Arrow combinator in a first version of the ArrowParallel typeclass}
\label{fig:parEvalNArrowTypeClass1}
\end{figure}
Sometimes parallel Haskells require or allow for additional configuration parameters, \eg an information about the execution environment or the level of evaluation (weak-head normalform vs. normalform). For this reason we also introduce an additional \code{conf} parameter to the function. We also do not want \code{conf} to be a fixed type, as the configuration parameters can differ for different instances of \code{ArrowParallel}. So we add it to the type signature of the typeclass as well and get \code{ArrowParallel arr a b conf} (Fig.~\ref{fig:parEvalNArrowTypeClassFinal}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
class Arrow arr => ArrowParallel arr a b conf where
	parEvalN :: conf -> [arr a b] -> arr [a] [b]
\end{lstlisting}
\caption{parEvalN Arrow combinator in the final version of the ArrowParallel typeclass}
\label{fig:parEvalNArrowTypeClassFinal}
\end{figure}
Note that we don't require the \code{conf} parameter in every implementation. If it is not needed, we usually just default the \code{conf} type parameter to () and even blank it out in the parameter list of the implemented \code{parEvalN}, as we will see in the implementation of the Multicore and the ParMonad backend.
